---
title: "monkey_descirptive stats"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ook back at your monkey data (here I WILL ADD MORE THINGS

so after getting your PhD onthe new speices you are gonna do a postdoc to study them even more (yay!)

it is wide, lets make it long

```{r pivot}

```

what else can we do?

summary stats?

We need the mean of the taisl...

SSSSSSSS

sssss

lets do it the long way

## descriptive stats

```{r}
total <- sum(new_data$tail_length)
len <- length(new_data$tail_length)
total/len
```

lets make our first function

```{r}
find_mean <- function(data){
  a <- sum(data)
  b <- a/length(data)
  return(b)
}
find_mean(new_data$tail_length)
```

hey, we just made our firt function!!!!

(of coure, R has a buitl in mean fucito..but dont tell your prof that..let her think you are just asweomd!

but here is a trciky thing.we have a sample mean not the populaiton mean. that is cause this is just an esiamte to fhte true mean.

your prof was smart when she asked for 500 monkeys. the more smaples the closer your sample reaches to the population.

here is an inteartie veriosn that shows this well (doesnt work)

```{r}
library(emdbook)
library(plotly)
library(here)

m <- list(
  l = 50,
  r = 50,
  b = 10,
  t = 10,
  pad = 4
)
accumulate_by <- function(dat, var) {
  var <- lazyeval::f_eval(var, dat)
  lvls <- plotly:::getLevels(var)
  dats <- lapply(seq_along(lvls), function(x) {
    cbind(dat[var %in% lvls[seq(1, x)], ], frame = lvls[[x]])
  })
  dplyr::bind_rows(dats)
}

d <- do.call(rbind, lapply(lseq(10,10000, 300), function(x){
	d <- data.frame(x=rnorm(x), frame=x/300, N=x)
	return(d)
}))

dd <- aggregate(data=d, x~frame+N, FUN=mean)
%>% accumulate_by(~N)

p <- dd %>% plot_ly(x=~log10(N), y=~x, frame=~frame, type="scatter", mode="lines", line = list(simplyfy = F, color="orangered"),width=550, height=350)%>% 
 animation_opts(
    frame = 10, 
    transition = 0, 
    redraw = FALSE
 ) %>% 
config(displayModeBar = F) %>%
  layout(
    xaxis = list(
      title = "Sample Size (log10)",
      zeroline = F
    ),
    yaxis = list(range=c(-0.7,0.7),
      title = "Mean",
      zeroline = F
    ), autosize=F, margin=m
  ) %>%
  animation_slider(
    hide = T
    )%>%
  animation_button(
    x = 1, xanchor = "right", y = 0, yanchor = "bottom"
  )
```

[\<https://tinystats.github.io/teacups-giraffes-and-statistics/04_variance.html\>](https://tinystats.github.io/teacups-giraffes-and-statistics/04_variance.html){.uri}

[\<https://github.com/tinystats/teacups-giraffes-and-statistics/blob/master/04_variance.Rmd\>](https://github.com/tinystats/teacups-giraffes-and-statistics/blob/master/04_variance.Rmd){.uri}

## spread

variance

strat with popualtoi var

## diet

ok, you now need info on diet so you get info on how much food they eat and their body size.

we want to look at the covariance
