---
title: "visual_answers"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Answers

```{r question  1}

monkey_data_1 %>%  ggplot(aes(tail_length)) + geom_histogram(binwidth =  1,  fill="blue")

```

1.  put the legend on the bottom rather than on the right
2.  label the x axis "tail length (cm)"
3.  label the y axis "number of observations"
4.  give it a title that is meaningful
5.  put a credit somewhere that says who made the figure
6.  change the legend so that it says "Monkey groups"
7.  change the labels from "a" and "b" to "Group one" and "Group 2"
8.  make a table to have at the back

```{r  question 2: making the graph}
all_data %>% ggplot(aes(tail_length, fill=monkey_pop )) + geom_histogram(color="#e9ecef", alpha=0.6, position = 'identity') + theme(legend.position = "bottom") + labs(title = "monkey's from Naboombu", caption = "data collected by Jessica Jones", x= "tail lenght (cm)", y ="number of monkeys", fill = "groups") + scale_fill_discrete(breaks=c("a", "b"),
                         labels=c("Group a", "group b"))
```

## [New visualizations]{.ul}

```{r question 3}
ll_data %>% ggplot(aes(tail_length, fill=monkey_pop )) + geom_density(color="#e9ecef", alpha=0.6, position = 'identity')


library(ggridges)

ggplot(all_data, aes(x = tail_length, y = monkey_pop)) + geom_density_ridges()

ggplot(new_data, aes(x = tail_length, y = monkey_pop, fill= monkey_pop)) + geom_density_ridges()

ggplot(new_data, aes(x = tail_length, y = monkey_pop, fill= monkey_pop)) + geom_density_ridges(aes(alpha = .3))


### box

all_data %>%   ggplot(aes(tail_length, monkey_pop, fill = monkey_pop)) + geom_boxplot()

all_data %>%   ggplot(aes( monkey_pop, tail_length, fill = monkey_pop)) + geom_boxplot()

#violin

all_data %>%   ggplot(aes( monkey_pop, tail_length, fill = monkey_pop)) + geom_violin()



all_data %>%   ggplot(aes(monkey_pop, tail_length, fill = monkey_pop)) + geom_violin()+ geom_jitter()
```
