---
title: "some notes on working with data tables in R"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```






```{r}
library(tidyverse)
```


```{r}
parse_integer(c("123", "345", "abc", "123.45"))


parse_number(c("123", "345", "abc", "123.45"))

```

#parsing numbers from data is tricky
# People write numbers differently in different parts of the world. 
#For example, some countries use . and other , before the fraction
# 
# Numbers are often surrounded by other characters that provide some context, like “$1000” or “10%”.
# 
# Numbers often contain “grouping” characters to make them easier to read, like “1,000,000”, and these grouping characters vary around the world.

Example of how to fix

```{r}
parse_double("1.23")
parse_double("1,23")

parse_double("1,23", locale = locale(decimal_mark = ","))


# also, parse number just ignores most other things

parse_number("the gift cost $123.45")

```


```{r}
charToRaw("Marc")
#hexadecmial
```

the mapping from hexadecimal number to character is called the encoding. here  encoding is called ASCII. ASCII does a great job of representing English characters, because it’s the American Standard Code for Information Interchange.
BUt this can fail if we have data that isn't in English  Characters

to fix this, one common encoding used is called UTF-8 (uber fun fact: this lets you do emojis!)

because of all of this, readr uses and assumes UTF-8 encoding

but sometimes this is bad if data is from an older system and it gives back werid stuff

```{r}

x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd"

parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

Good datasets tell us the encoding used but that is also not that common. Readr will guess the encoding for you but this sometimes fails or isn't ideal

```{r}
guess_encoding(charToRaw(x2))
```


# dates etc


parse_datetime() expects an ISO8601 date-time.
ISO 8601 date  organised from biggest to smallest: year, month, day, hour, minute, second.


```{r}
parse_datetime("2010-10-01T2010")

parse_datetime("20101010")
#no time so it sets it to midnight
```

parse_date() expects a four digit year, a - or /, the month, a - or /, then the day:


```{r}
parse_date("2010-10-01")
```


Year
%Y (4 digits).
%y (2 digits); 00-69 -> 2000-2069, 70-99 -> 1970-1999.

Month
%m (2 digits).
%b (abbreviated name, like “Jan”).
%B (full name, “January”).

Day
%d (2 digits).
%e (optional leading space)

Non-digits
%. skips one non-digit character.
%* skips any number of non-digits.



```{r}
parse_date("01/02/15", "%m/%d/%y")



parse_date("01/02/15", "%d/%m/%y")

parse_date("01/02/15", "%y/%m/%d")
#> [1] "2001-02-15"```
```


Non-digits
%. skips one non-digit character.
%* skips any number of non-digits.


# readr
```{r}
challenge <- read_csv(readr_example("challenge.csv"))
problems(challenge)
```


```{r}

challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_logical()
  )
)


challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
challenge


```


# separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears. Take table3:



```{r}
table3
```


```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

By default, separate() will split values wherever it sees a non-alphanumeric character (i.e. a character that isn’t a number or letter). For example, in the code above, separate() split the values of rate at the forward slash characters. If you wish to use a specific character to separate a column, you can pass the character to the sep argument of separate(). For example, we could rewrite the code above as:


```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```


```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```



```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```



# unite


combines multiple columns into a single column. 

```{r}
table5
```


```{r}
table5 %>% 
  unite(full_year, century, year)
```


# missing values

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```


The return for the fourth quarter of 2015 is explicitly missing, because the cell where its value should be instead contains NA.

The return for the first quarter of 2016 is implicitly missing, because it simply does not appear in the dataset.

**An explicit missing value is the presence of an absence; an implicit missing value is the absence of a presence.**


```{r}
stocks %>% 
  complete(year, qtr)
```

sometimes folks use missing values to indicate that the above value should be filled in

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

```

