---
title: "R live- day 1"
author: "Marc Kissel"
date: "Feb 1, 2021"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro section

## [Why use R]{.ul}

it is more than just a way to run statistics. You can make nice figures,
create documents like this one, run computer simulations, create
interactive apps, and even [make a
website](https://marckissel.netlify.app/ "my site")!

I don't know where i read this analogy, but it helps to think of R
compared to SPSS, Excel, etc as looking at the difference between taking
a bus and driving a car. Taking a bus is easy and carefree & it gets you
from point A to point B. But there is not much freedom in where you go.
With a car you can go anywhere, but you have to learn how to drive and
pay attention during the trip. In R you can do almost anything with your
data. But it takes some time and effort...though i think it is worth it!

## [How to get help]{.ul}

No one knows everything about R. I often Google the same issue many
times since i forget. A great place to look for help is [Stack
Overflow](https://stackoverflow.com/). If you are on Twitter, lots of
great folks follow the
[\#Rstats](https://twitter.com/search?q=%23rstats). Often, I find the
best way to Google it is to take the thing you want to know and add
"rstats" to the end or the name of the package i'm using. So, if i
wanted to change the background color of a plot "ggplot2 background
color" brings up a link to a stackoverflow discussion on it.

Also, most packages have help online. Some even have tutorials or
vignettes. I like to run through the online tutorial of a new package
since it helps me to figure out what is going on.

A short warning: if you ask a question online be prepared to give
details on your question. It might be better to ask me first and if i
don't know the answer i can help you get your question into a format
that allows others to help you!

# [Working with R]{.ul}

The *console* is where commands are run in RStudio. You can write
something in the console to speak directly to R, or write in the
*source* and then send the script to the console. You can recognize the
console since every lines starts with a '\>'

go to the console and type the following and then hit enter

```{r}
7 + 6
```

you should see something like this:

> 7 + 6 [1] 13

note the [1], which tells you the line number. This becomes useful later
on. Also note it told you that 7 + 6 = 13.

## [Short overview of how R works]{.ul}

One way that i think about R that helps me to work it is to remember
that everything in R is an "object." What this means is that in R we
make objects (or we assign values to objects) and then do things to
these objects. for example check out the code below

```{r assign value}
a <- 4

```

What this says is that we are going to create the object '**a**' and
then assign the value '4' to it. If you run the code in R Studio you
will see that the 'environment' panel now lists 'a' and gives its value.

> **The arrow shortcut key for a pc is "alt" + "-"". For a mac it is
> "Option" + "-"**

```{r objects}
b <- "my name"
c <- FALSE
d <- c(1,2,3.0,4)
```

> **protip:** When you assign values to objects, be aware that not all
> names are possible. R reserves some names for special functions. here
> are the rules
>
> 1.  Can use letters/numbers, but the object has to start with letter
>     (impt for when you import data).
> 2.  no spaces, so use CamelCase or Snake_case

[**to store more than one element in an object we need to remember a
trick**]{.ul}

if you want to make a series of numbers, characters, whatever, you need
to use a special sequence. To do this, we must combine the values we
want to assign with the c function, which combines values into a vector!

```{r}
my_list <- c(1,2,3,4,5)
my_list_2 <- c("my_name", "my_address", "my_number")
```

## [Using objects]{.ul}

Once you assign *something* (e.g., a value, a list of numbers, a
character vector) to a object you can interact with that object. We do
this by using what is called a **function**. A function is a built-in R
command that does things to objects. Functions take what we call
**arguments**.

To call a function in R you write the name of the function and then put
the arguments of that function in parenthesis. Each argument is
separated by a comma. Some functions have one argument but many have
more. Often, the arguments have default values which means you don't
have to specify what that is. For example the **mean()** function
defaults to na.rm = False, which means it won't automatically remove
missing values. Oftentimes you might want to skip the missing values, so
you would add na.rm = TRUE

R read "NA" as standing for missing values. However, it is counted as an
entry. so if you have NA values you need to remember to tell R to skip
those if calculating the mean

```{r functions}
sum(d)
mean(d)
mean(d, na.rm = FALSE) #doesn't change result since we have no NAs yet
e <- c(1,2,3,4,NA,5)
mean(e)
mean(e, na.rm = TRUE)
seq(from = 1, to = 10, by =2)

```

if you need help on a function, you can type a question mark before its
name like this: ?mean()

The way in which you can interact with the object is based on the
*class* of the object. The class of the object is assigned by R based on
what it thinks the object is supposed to be.

```{r}
class(a)
class(b)
class(c)
class(d)
```

In this case, the function is 'class' and the argument is the object
whose class you want to know.

## [Class]{.ul}

A **class** defines what kinds of operations can be implemented on an
object & how a function will return a value. It is important to keep
track of the classes of your objects. Class mistakes are probably the
most common kind of problem in

Once you have an object stored, you can interact with it but remember
that the way you interact with it via functions depends on its class....

```{r}
length(my_list)
length(my_list_2)
sum(my_list)
my_list[4] #gets the 4th value.
#sum(my_list_2) this kicks back an error
```

running the second line gives you a message like this:

"Error in sum(my_list_2) : invalid 'type' (character) of argument"

What this means is that the 'type' or 'class' of the object is not
something that the sum function knows what to do with.

[**Learning how to read R errors is useful.**]{.ul}

If you don't know what the error means, copying part of it and searching
Google is a pretty good way to find help.

(*short digression: one of the things about R that makes it easier to
use is that many of the functions work by first checking out the class
of the object you are asking it to work on. What this means is that the
same function can be applied to different types of classes. This doesn't
really affect much now, but as you investigate R more you can see how
this works*.)

## [Starting a project]{.ul}

1.  best advice i can give is that you want everything to be in one
    place (makes things a bit harder at first)

2.  this is what R Projects are for

3.  start a new project and store all your data there (see handout from
    Week 1)

4.  when you start/open a project, it sets R's working directory to that
    folder. This way, you can easily locate things

## [Packages]{.ul}

R comes with many functions already installed. But what makes it super
powerful is that people can add to these functions by writing their own
(and we will write our own function in week 7!). A bunch of functions
bundled together is called a **Package**. You are going to find lot of
useful packages as you us R. To install a package in R we use the
function "install.packages," which has the argument the name of the
package.

install.packages("name_of_package") \#install a package
library(name_of_package) \#load the package for your session

```{r}
#install.packages("tidyverse")

```

you only have to install a package once, but [**everytime you reopen R
you have to reload the pacakage with the library function**]{.ul}

```{r}
library(tidyverse)
```

In the case of this package, R tells us what is being loaded and also
that there are some conflicts. We can usually ignore these, but
basically it just says that some functions from this new package have
the same name as functions from the base R package. Good to pay
attention to but usually not a huge problem unless you are doing a lot
of things in one script

Here is a short piece of code that might make some sense. It calls a
function ggplot (part of the tidyverse package), and then makes a plot.
Then the ggsave function saves the new plot as a pdf in the working
directory of the project folder.

the last line takes the data set that was used (diamonds) and write as
CSV file for that.

```{r}
ggplot(diamonds, aes(carat, price)) + 
  geom_point()
ggsave("diamondPlot.pdf")
write_csv(diamonds, "diamonds3.csv")
```

# [How to get data into R]{.ul} 

Probably the hardest part of starting to use R is getting your data into
R and then being able to work with it fluidly. This has gotten easier
over time but still can be tricky. For me, the biggest hurdles are

1.  getting the file in a good format (which is next week's work!)

2.  making sure that when it is read into R all of the data is being
    read correctly by R

    ## Using readr

The readr package is part of the tidyverse package. One of the points of
the tidyverse is to try to make things consistent.

Here are some of the functions to read data into R

read_csv - comma separated values

read_csv2() - semicolons

read_fwf() - fixed width

For all of these the first argument is the file name. Other arguments
can be used to if need be. For example, use "col_names = FALSE" if no
headers to your data (uncommon)

What is nice about read_csv etc is that it tells you what it **thinks**
that class of each column is

For more info check this out:

<https://readr.tidyverse.org/> or
<http://r4ds.had.co.nz/data-import.html>

You can also use the 'import data set' tool which lets you 'point and
click' and then generates the code. This is helpful and probably
something i should use more often

### Things to be aware of when importing data into R

As we talked about, the biggest problem is dealing with factors. factors
are categorical data. In the old days R used to import any string as a
factor. This caused some problems! Now, readr fixes this. But we still
need to be aware of how factors work. The following code, taken mostly
from <http://www.ats.ucla.edu/stat/r/modules/factor_variables.htm> gives
a workthough on what factors are and how to think about them.

We are going to start by making a new object and then turning it into a
factor. then play around with this new factor to see how it works

\#\#\#Factors

```{r factors}



set.seed(100) #this sets the randomizer so if you run this code you get the same results i did
temp <- sample(0:1, 20, replace=T)
temp

is.factor(temp)
is.numeric(temp)
#ok, so we now have a list of numbers
#remember that factors are categories
#so, we are going to assign factors to these values
#uses the argument labels
temp_f <- factor(temp, labels = c("femur", "tibia"))
temp_f #hey, that is kinda cool! lets check if it still numeric
is.numeric(temp_f)
is.factor(temp_f) #sweet!
#so that works with numbers, but what about strings

size <- c("small", "medium", "small", "small", "small", "small", "medium", "small",
          "medium", "medium", "medium", "medium", "medium", "tall", "tall",
          "small", "medium", "medium", "small", "tall")
is.factor(size)
is.character(size)
#ok, so how do we turn this into a factor? lets what we did before
size_f <- factor(size)
is.factor(size_f)
size_f #but lets take a closer look at the levels..
?levels #read the help here and try to see what it does
levels(size_f)
# ok, so notice the order of the levels. here, they are sorted alphabetically
#this might not be what you want (in fact, it rarely is.). I've had many
#issues with R that comes down to this problem. Dplyr/ggplot makes it a little
#easier, but you still going to have problems
#so, how to fix this:

#we need to set the levels!
size_f2 <- factor(size, levels=c("small", "medium", "tall"))

#the above code says somthing like "take the size vector and set the levels so that are in this order: small, medium, tall

is.factor(size_f2)
levels(size_f2) #woot. they are now in the right 

###
#sometimes you might need to use ordinal data. for that we use ordered factors
#note: ordinal data is when we the order of the values matters,
#but we can't say how much of a difference there is between each.
#the data is ranked but the distance between categories isn't clear
#so, if we are measuring something like job satisifaction
#hard to say how much better 'very happy' is from 'happy'
#e.g. likert scale

size_ordered <-  ordered(size, levels =c("small", "medium", "tall"))
size_ordered
size_f2
#note in size_ordered you see the '<' 

table(size)
table(size_ordered)

#the last thing i want to mention is adding data. now,
#you might think this is easier to do in excel
#but remember that R has the plus of perserving
#all you steps. Plus, this way you aren't messing with
#the original data

size_f2[21] <- "very.tall"
#we get an error message. note it says 'invalid factor'
size_f2 
#there is now a NA. this is cause R doesn't know how to deal 
#with the new factor
#we need to add the new factor
size_f2 <- factor(size_f2, levels = c(levels(size_f2), "very.tall")) #ok, so this
#is a bit long. but levels is the same. just cheating a bit by usinf the levels(size_f2)
#function to get the original levels from the list. As you getbetter w/ R
#you learn these little tricks to help. avoids making mistakes...
size_f2
size_f2[21] <- "very.tall"
####ok, but lets say we don't want 'very tall' anymore. This could become
#problem when making images since it may still show up even if we just
#remove it
#first, we get rid of the elements we dont want
size_f2_new <- size_f2[size_f2 != "very.tall"]
#then, we 'refactor' it
size_f2_new <- factor(size_f2_new)
size_f2_new
#note the difference between the 2 tables
table(size, temp)
table(size_f2_new, temp_f)


#finally, lets put it all together
data <- c(34, 39, 63, 44, 47, 47, 57, 39, 48, 47, 34, 37, 47, 47, 39, 47, 47, 50, 28, 60)

my_data <- data.frame(temp, temp_f, size, size_f2_new, data)
my_data
```

That was a bit much, i know. It took me a long time to get a handle on
this stuff. Factors are notorious for being the worst part of R. Come
back to this section later in the class when you are having issues and
hopefully it makes more sense then!

### F[actors in the tidyverse (see <http://r4ds.had.co.nz/factors.html> for more on working with factors in R)]{.ul}

#### 1. reorder factors.

Often when making plots we want to change the order they appear in. One
was is shown above. Another way is to use the forcats package that comes
with the tidyverse. it comes with an interesting dataset that, among
other things, tell us the avg daily hours someone watches tv based on
their religion

lets set the data up:

```{r}
library(tidyverse) # be sure to load the package
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(x = tvhours,  y= relig)) + geom_point() # this function ggplot takes a lot fo arguments and we will work on it soon. for now note that we give it the X and Y axis and then tell it to make a point graph 
```

when we plot this the factors are not organized by tv hours, which makes
it hard to examine.

We can improve it by reordering the levels of relig using fct_reorder().

fct_reorder() takes 2 main arguments:

1.  the factor we want to reorder

2.  the numeric vector we want to reorder by

```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) + 
  geom_point()
```

#### rename factors

In the Tidyverse it is easy to rename factors using the function
fct_recode()

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
                              "Republican, strong"    = "Strong republican",
                              "Republican, weak"      = "Not str republican",
                              "Independent, near rep" = "Ind,near rep",
                              "Independent, near dem" = "Ind,near dem",
                              "Democrat, weak"        = "Not str democrat",
                              "Democrat, strong"      = "Strong democrat"
  )) %>%   count(partyid)
```

# [More on the Tidyverse]{.ul}

Next week i'll do more on this. For now this should help you with the
homework and getting ready to work with R.
